Validation of Visual Statistical Inference, Applied to Linear Models (supplementary materials)

### 2013-04-16 paper_jasa.tex
supplementary materials and reference updated

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. The lineup protocol enables statistical significance testing of visual findings, bridging the  gulf between exploratory and inferential statistics.  In this paper inferential methods for statistical graphics are developed further by refining the terminology of visual inference, and framing the lineup protocol in a context that allows direct comparison with conventional tests in scenarios when a conventional test exists. This framework is used to compare the performance of the lineup protocol against conventional statistical testing in the scenario of fitting linear models. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol performs comparably with the conventional tests, and expectedly out-performs them when data is contaminated, a scenario where assumptions required for performing a conventional test are violated. Surprisingly, visual tests have higher power than the conventional tests when the effect size is large. And, interestingly, there may be some super-visual individuals who yield better performance and power than the conventional test even in the most difficult tasks.

@@

An illustration of the lineup protocol in relation to conventional hypothesis testing is presented in Table \ref{tbl:compare}. Both methods start from the same place, the same set of hypotheses. The conventional test statistic is the $t$-statistic, where the parameter estimate is divided by its standard error. In the lineup protocol, the test statistic is a plot of the data. Here, side-by-side boxplots are used, because the variable of interest is categorical and takes just two values. In conventional hypothesis testing the value of the test statistic is compared with all possible values of the sampling distribution, the distribution of the statistic if the null hypothesis is true. If it is extreme on this scale then the null hypothesis is rejected. In contrast in visual inference, the plot of the data is compared with a set of plots of  samples drawn from the null distribution. If the actual data plot is selected as the most different, then this results in rejection of the null hypothesis.

@@

\noindent with $X$ defined as above, and $x$ is the number of observers selecting the actual data plot. 
\end{dfn}
\noindent Note that for $x=0$ the $p$-value becomes, mathematically, equal to 1. It might make more sense from a practical point of view to think of the $p$-value as
being larger than $P(X \ge 1)$ in this situation. By increasing either $m$ or $K$, the value at a higher precision can be determined. Table \ref{pvalue} shows $p$-values for different numbers of observers for lineups of size $m = 20$. %It can be seen that, if the null hypothesis is true, it is unlikely that multiple observers would pick the actual data plot.

@@

\begin{dfn}\label{dfn:visualtest}
The \textbf{visual test}, $V_{\theta}$  of  size $m$ and significance level $\alpha$, is defined to  
\begin{itemize}\itemsep-10pt
\item \textbf{Reject} $H_0$ if out of $K$ observers at least $x_{\alpha}$ correctly identify the actual data plot, and
\item \textbf{Fail to reject} $H_0$  otherwise. 
\end{itemize}
where $x_{\alpha}$ is such that $P(X \ge x_{\alpha}|H_0) \le \alpha$. 
\end{dfn}

@@

\noindent Because $X$ takes only discrete values we can not always control exactly for $\alpha$.  
For example, when there is only one observer, $1/m$ is the minimal value at which we can set $\alpha$. It can be set to be smaller, even arbitrarily small, by increasing $K$, the number of observers. 
Type II error is harder to calculate, as is usually the case. In visual inference, individual abilities need to be incorporated to calculate Type II error. %This is not necessary for Type I error, because it is simply the chance that the observer picks the actual data plot, by chance. 
Here, we need to estimate the probability that an observer sees the actual data plot as different, when it really is different. This involves understanding the individual's visual skills. Thus, let $X_i$ be a binary random variable with $X_i = 1$, if individual $i \ (=1, \dots , K)$ identifies the actual data plot from the lineup, and $X_i = 0$ otherwise. Let $p_i$ be the probability that individual~$i$ picks out the actual data plot. If all individuals have the same ability, with the probability, $p$, for picking out the actual data plot, then $X = \sum_i X_i$ has distribution $\text{Binom}_{K, p}$, and we can estimate $p$ by $\hat{p} = x/K$, where $x$ is the number of observers (out of $K$), who pick out the actual data plot.

@@

\begin{proof} The proof and the details of the lemma are attached in the supplementary documents. 
\end{proof}

@@

In this setting, there are many established graphics that are used to evaluate and diagnose the fit of a regression model (e.g. \citet{cook:99}). Table \ref{tbl:stat_multiple} lists several common hypotheses related to the regression setting, and commonly used statistical plots that might be used as corresponding visual test statistics.
%These plots should be familiar, because they are all in common use. 
For example, to examine the effect of variable $X_j$ on $Y$, we would plot residuals obtained from fitting the model without $X_j$ against $X_j$ or for a single covariate we may plot $Y$ against $X_j$ (cases 1-4 in Table~\ref{tbl:stat_multiple}). To assess whether the linear assumption is appropriate we would draw a plot of residuals against fitted values (case 5 in Table~\ref{tbl:stat_multiple}). For the purpose of comparing visual against conventional inference, we focus on cases 2 and 3, with a continuous and categorical explanatory variable, respectively.

@@

\begin{table*}[hbtp] 
\caption{Visual test statistics for testing hypotheses related to the model $Y_i = \beta_0 + \beta_1 X_{i1} + \beta_1 X_{i2} + \beta_3 X_{i1}X_{i2} + ... + \epsilon_i  $ } 
\centering 
\begin{tabular}{m{0.5cm}m{3cm}m{2cm}m{3cm}m{5.5cm}} 
\hline\hline 
Case & Null Hypothesis & Statistic & Test Statistic & Description \\ [0.5ex] % inserts table %heading 
\hline 
1 & $H_0: \beta_0=0$ & Scatter plot & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_intercept.pdf}}\end{minipage} & Scatter plot with least square line overlaid. For null plots   we simulate data from fitted null model. \\ 
2 & $H_0: \beta_k=0$ & Residual plot & \begin{minipage}[t]{3cm}   \scalebox{0.4}{\includegraphics{stat_beta_k.pdf}}\end{minipage} & Residual vs $X_k$ plots. For null plots   we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
3 & $H_0: \beta_k=0$ (for binary $X_k$) & Box plot & \begin{minipage}[t]{3cm} 	\scalebox{0.4}{\includegraphics{stat_category.pdf}} \end{minipage} & Box plot of residuals grouped by category of $X_k$. For null plots   we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\
4 & $H_0: \beta_k=0$ (interaction of continuous and  binary $X_k$) & Scatter plot & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_interection.pdf}}  \end{minipage} & Scatter plot with least square lines of each category overlaid. For null plots   we simulate data from fitted null model.\\[1ex] % [1ex] adds vertical space 
5 & $H_0: X$ Linear & Residual Plot & \begin{minipage}[t]{3cm}  	\scalebox{0.4}{\includegraphics{stat_nonlinear.pdf}}  \end{minipage} & Residual vs predictor plots with loess smoother overlaid. For null plots   we simulate residual data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
6 & $H_0: \sigma^2=\sigma^2_0$ & Box plot & \begin{minipage}[t]{3cm} \scalebox{0.4}{\includegraphics{stat_sigma_box.pdf}}\end{minipage} & Box plot of standardized residual divided by  $\sigma^2_0$. For null plots   we simulate data from standard normal. \\ 				
7 & $H_0: \rho_{X,Y|Z}=\rho$ & Scatter Plot & \begin{minipage}[t]{3cm}  	\scalebox{0.4}{\includegraphics{stat_intercept.pdf}}  \end{minipage} & Scatter plot of Residuals obtained by fitting partial regression. For null plots   we simulate data (mean 0 and variance 1) with specific correlation $\rho$. \\ 
8 & $H_0:$ Model Fits & Histogram & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_goodness_simple.pdf}}  \end{minipage} & Histogram of the response data. For null plots   we simulate data from fitted model. \\[1ex] % [1ex] adds vertical space 
9 & Special case $p=1$ $H_0: \rho_{X,Y} =\rho$ & Scatter plot & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_intercept.pdf}}  \end{minipage} & Scatter plot with least square line overlaid For null plots   we simulate data with correlation $\rho$.\\			
\hline 
\end{tabular} 
\label{tbl:stat_multiple} 
\end{table*}

@@

\begin{equation}\label{power_exp} 
%   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
 p(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test for different $K$ (number of observers), obtained using $p(\beta)$ from Equation \ref{power_exp}. Notice that the expected power of the visual test exceeds the power of the conventional test as $K$  increases, and when $\beta$ gets larger.   Conversely, visual power is below conventional power for parameter values close to the null hypothesis. This is even more pronounced for large number of observers. At the same time, the point of intersection between visual and conventional power approaches the value of the null hypothesis as the number of observers approaches infinity, leading to an asymptotically perfect power curve of zero in the null hypothesis and one for any alternative value. We observe this dichotomy of visual power  in the data collected from user experiments, too. It features prominently in  figure \ref{fig:power_loess_effect}.

@@

Three experiments were conducted to evaluate the effectiveness of the lineup protocol relative to the equivalent test statistic used in the regression setting. The first two experiments have  ideal scenarios for conventional testing, where we would not expect the lineup protocol to do better than the conventional test. The third experiment is a scenario where assumptions required for the conventional test are violated, and we would expect the lineup protocol to outperform the conventional test. (Data and lineups used in the experiments are available in the supplementary material.)

@@

\subsection{Data Cleaning}

@@

\subsection{How much do null plots affect the choice?}

@@

\subsection{Type III error}~\label{sec:TypeIII}

@@

This paper has demonstrated that statistical graphics can be used in statistical inference and validates the lineup protocol proposed by \citet{buja:2009}. Specific terminologies were defined, and methods for obtaining the $p$-value and estimating the power of visual tests were introduced. It was assumed that observers select the plot having the strongest signal in the lineup, in order to calculate the theoretical power,  and the experimental data was consistent with this assumption.  The simulation study showed that the results obtained by visual inference are comparable to those obtained by conventional inference, in the setting where conventional testing can be done. Visual inference is intended to provide valid tests where no conventional test exists, and our experiments in a controlled scenario suggest that it will perform as expected in the intended applications. The power of a visual test increases with the number of observers, which interestingly, leads to a result that the theoretical power of visual test can be better than that of conventional test.

The lineup protocol operates similarly to statistical tests that have broad alternative hypotheses. If the null hypothesis is rejected, generally we can say that ``there is something there'' but not specifically what it is in the data that triggers the rejection. Follow-up questions on the reasons provide qualitative insight. In conventional testing, multiple comparisons are often done to refine and understand the test results, and perhaps some similar approaches might be developed for visual inference.

@@

Visual inference has been successfully used in two practical applications: to evaluate the power of competing graphical designs \citep{heike:2012}, and to detecting signal presence in large $p$, small $n$ data \citep{niladri:2012}. It is hoped that the lineup protocol will prove to be  valuable in data mining applications, and exploratory analyses, where there are no existing gauges of statistical significance.

@@

%\bibliographystyle{plain}
%\bibliographystyle{plainnat}
\bibliographystyle{asa}
%\bibliographystyle{ieeetr}
\bibliography{references}

>>>>

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. A new method, called the lineup protocol, enables statistical significance testing of visual findings, which bridges a big gulf between exploratory and inferential statistics.  In this paper inferential methods for statistical graphics are developed further by refining the terminology of visual inference, and framing the lineup protocol in a context that allows direct comparison with conventional tests, in scenarios when a conventional test exists. This framework is used to compare the performance of the lineup protocol against conventional statistical testing in the scenario of fitting linear models. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol performs comparably with the conventional tests, and expectedly out-performs them when data is contaminated, a scenario where assumptions required for performing a conventional test are violated. Surprisingly, visual tests have higher power than the conventional tests when the effect size is large. And, interestingly, there may be some super-visual individuals who yield better performance and power than the conventional test even in the most difficult tasks.

@@

An illustration of the lineup protocol in relation to conventional hypothesis testing is presented in Table \ref{tbl:compare}. Both methods start from the same place, the same set of hypotheses. The conventional test statistic is the $t$-statistic, where the parameter estimate is divided by its standard error. In the lineup protocol, the test statistic is a plot of the data. Here, side-by-side boxplots are used, because the variable of interest is categorical and takes just two values. In conventional hypothesis testing the value of the test statistic is compared with all possible values of the sampling distribution, the distribution of the statistic if the null hypothesis is true. If it is extreme on this scale then the null hypothesis is rejected. In contrast in visual inference, the plot of the data is compared with plots of a finite number of samples drawn from the null distribution. If the actual data plot is selected as the most different, then this results in rejection of the null hypothesis.

@@

\noindent with $X$ defined as above, and $x$ is the number of observers selecting the actual data plot. 
\end{dfn}
\noindent Note that for $x=0$ the $p$-value becomes, mathematically, equal to 1. It might make more sense from a practical point of view to think of the $p$-value as
being larger than $P(X \ge 1)$ in this situation. By increasing either $m$ or $K$, the value at a higher precision can be determined. Table \ref{pvalue} shows $p$-values for different numbers of observers for lineups of size $m = 20$. It can be seen that, if the null hypothesis is true, it is unlikely that multiple observers would pick the actual data plot.

@@

\begin{dfn}\label{dfn:test}
The \textbf{visual test}, $V_{\theta}$  of  size $m$ and significance level $\alpha$, is defined to  
\begin{itemize}\itemsep-10pt
\item \textbf{Reject} $H_0$ if out of $K$ observers at least $x_{\alpha}$ correctly identify the actual data plot, and
\item \textbf{Fail to reject} $H_0$  otherwise. 
\end{itemize}
where $x_{\alpha}$ is such that $P(X \ge x_{\alpha}|H_0) \le \alpha$. 
\end{dfn}

@@

\noindent Because $X$ takes only discrete values we can not always control exactly for $\alpha$.  
For example, when there is only one observer, $1/m$ is the minimal value at which we can set $\alpha$. It can be set to be smaller, even arbitrarily small, by increasing $K$, the number of observers. 
Type II error is harder to calculate, as is usually the case. In visual inference, individual abilities need to be incorporated to calculate Type II error. %This is not necessary for Type I error, because it is simply the chance that the observer picks the actual data plot, by chance. 
Here, we need to estimate the probability that an observer sees the actual data plot as different, when it really is different. This involves understanding the individual's visual skills. Thus, let $X_i$ be a binary random variable with $X_i = 1$, if individual $i \ (=1, \dots , K)$ identifies the actual data plot from the lineup, and $X_i = 0$ otherwise. Let $p_i$ be the probability that individual $i$ picks out the actual data plot. If all individuals have the same ability, with the probability, $p$, for picking out the actual data plot, then $X = \sum_i X_i$ has distribution $\text{Binom}_{K, p}$, and we can estimate $p$ by $\hat{p} = x/K$, where $x$ is the number of observers (out of $K$), who pick out the actual data plot.

@@

\begin{proof} The proof and the details of the lemma are attached to the supplementary documents. 
\end{proof}

@@

In this setting, there are many established graphics that are used to evaluate and diagnose the fit of a regression model (e.g. \citet{cook:99}). Table \ref{tbl:stat_multiple} lists several common hypotheses related to the regression setting, and commonly used statistical plots that might be used as corresponding visual test statistics.
%These plots should be familiar, because they are all in common use. 
For example, to examine the effect of variable $X_j$ on $Y$, we would plot residual obtained from fitting model without $X_j$ against $X_j$ or for single covariate we may plot $Y$ against $X_j$ (cases 1-4). To assess whether the linear assumption is appropriate we would draw a plot of residuals against fitted values (case 5). For the purposes of comparing visual against conventional inference, we focus on cases 2 and 3, with a continuous and categorical explanatory variable respectively.

@@

\begin{table*}[hbtp] 
\caption{Visual test statistics for testing hypotheses related to the model $Y_i = \beta_0 + \beta_1 X_{i1} + \beta_1 X_{i2} + \beta_3 X_{i1}X_{i2} + ... + \epsilon_i  $ } 
\centering 
\begin{tabular}{m{0.5cm}m{3cm}m{2cm}m{3cm}m{5.5cm}} 
\hline\hline 
Case & Null Hypothesis & Statistic & Test Statistic & Description \\ [0.5ex] % inserts table %heading 
\hline 
1 & $H_0: \beta_0=0$ & Scatter plot & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_intercept.pdf}}\end{minipage} & Scatter plot with least square line overlaid. For lineup plot, we simulate data from fitted null model. \\ 
2 & $H_0: \beta_k=0$ & Residual plot & \begin{minipage}[t]{3cm}   \scalebox{0.4}{\includegraphics{stat_beta_k.pdf}}\end{minipage} & Residual vs $X_k$ plots. For lineup plot, we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
3 & $H_0: \beta_k=0$ (for categorical $X_k$) & Box plot & \begin{minipage}[t]{3cm} 	\scalebox{0.4}{\includegraphics{stat_category.pdf}} \end{minipage} & Box plot of residuals grouped by category of $X_k$. For lineup plot, we simulate data from normal with mean 0 variance $\hat{\sigma}^2$. \\
4 & $H_0: \beta_k=0$ (interaction with categorical $X_k$) & Scatter plot & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_interection.pdf}}  \end{minipage} & Scatter plot with least square lines of each category overlaid. For lineup plot, we simulate data from fitted null model.\\[1ex] % [1ex] adds vertical space 
5 & $H_0: X$ Linear & Residual Plot & \begin{minipage}[t]{3cm}  	\scalebox{0.4}{\includegraphics{stat_nonlinear.pdf}}  \end{minipage} & Residual vs predictor plots with loess smoother overlaid. For lineup plot, we simulate residual data from normal with mean 0 variance $\hat{\sigma}^2$. \\ 
6 & $H_0: \sigma^2=\sigma^2_0$ & Box plot & \begin{minipage}[t]{3cm} \scalebox{0.4}{\includegraphics{stat_sigma_box.pdf}}\end{minipage} & Box plot of standardized residual divided by  $\sigma^2_0$. For lineup plot, we simulate data from standard normal. \\ 				
7 & $H_0: \rho_{X,Y|Z}=\rho$ & Scatter Plot & \begin{minipage}[t]{3cm}  	\scalebox{0.4}{\includegraphics{stat_intercept.pdf}}  \end{minipage} & Scatter plot of Residuals obtained by fitting partial regression. For lineup plot, we simulate data (mean 0 and variance 1) with specific correlation $\rho$. \\ 
8 & $H_0:$ Model Fits & Histogram & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_goodness_simple.pdf}}  \end{minipage} & Histogram of the response data. For lineup plot, we simulate data from fitted model. \\[1ex] % [1ex] adds vertical space 
9 & Special case $p=1$ $H_0: \rho_{X,Y} =\rho$ & Scatter plot & \begin{minipage}[t]{3cm}  \scalebox{0.4}{\includegraphics{stat_intercept.pdf}}  \end{minipage} & Scatter plot with least square line overlaid For lineup plot, we simulate data with correlation $\rho$.\\			
\hline 
\end{tabular} 
\label{tbl:stat_multiple} 
\end{table*}

@@

\begin{equation}\label{power_exp} 
%   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
 p(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test for different $K$ (number of observers), obtained using $p(\beta)$ from Equation \ref{power_exp}. Notice that the expected power of the visual test exceeds the power of the conventional test as $K$ gets larger, and when $\beta$ gets larger.   Conversely, visual power is below conventional power for parameter values close to the null hypothesis. This is even more pronounced for large number of observers. At the same time, the point of intersection between visual and conventional power approaches the value of the null hypothesis as the number of observers approaches infinity, leading to an asymptotically perfect power curve of zero in the null hypothesis and one for any alternative value. This dichotomy of visual power can also be seen in more detail in figure \ref{fig:power_loess_effect}.

@@

Three experiments were conducted to evaluate the effectiveness of the lineup protocol relative to the equivalent test statistic used in the regression setting. The first two experiments have conventional ideal scenarios where we would not expect the lineup protocol to do better than the conventional test. The third experiment is a scenario where assumptions required for the conventional test are violated, and we would expect the lineup protocol to outperform the conventional test. (Data and lineups used in the experiments are available in the supplementary material.)

@@

\subsection{Data Cleaning}  \label{sec:data_cleaning}

@@

\subsection{How much do null plots affect the choice?}\label{sec:null_choice}

@@

\subsection{Type III error}~\label{sec:TypeIII}\label{sec:typeIII_error}

@@

This paper has demonstrated that statistical graphics can be used in statistical inference and validates the lineup protocol proposed by \citet{buja:2009} as a formal framework to accomplish that. For this, specific terminologies were defined. The methods of obtaining $p$-value and estimating power of visual tests were introduced. If it is assumed that observers tend to pick the plot with the strongest signal, which was consistent with the experimental data, the power of visual test can be obtained theoretically.  Interestingly, the theoretical power of visual test is better than that of conventional test in the conditions tested in this paper. The simulation study showed that even in the situation where the conventional test has its assumptions met, visual inference is comparable. The results also suggest that when the underlying assumptions of a conventional test do not hold, visual inference will perform well. Even though a single observer can evaluate a lineup, as suggested in the initial lineup protocol, this work recommends having multiple observers to ensure enough power.

When there is no conventional test available, visual testing procedure can be used due to its few assumptions and non-parametric nature.  Lineups are successfully used to evaluate the power of competing graphical designs in \cite{heike:2012}. This method is also applied in \citet{niladri:2012}. It is hoped that this proves to be extremely valuable in data mining applications, and exploratory analyses, where there are no existing gauges of statistical significance. The lineup protocol operates similarly to general tests, such as two-sided, analysis of variance $F$-tests, and Wilks $\Lambda$ for multivariate analysis of variance. If the null hypothesis is rejected, generally we can say that ``there is something there'' but not specifically what it is in the data that triggers the rejection. Follow-up questions on the reasons provide qualitative insight. Multiple comparisons are often done to refine and understand the results of the general conventional testing, and perhaps some similar approaches might be developed for visual inference.

@@

@@

%\bibliographystyle{plain}
%\bibliographystyle{plainnat}
%\bibliographystyle{ieeetr}
\bibliographystyle{asa}
\bibliography{references}

### 2013-04-16 supplementary.tex
supplementary materials and reference updated

@@

@@

\begin{table}[hbtp]
%\renewcommand\thetable{}
\caption{Selection of 10 lineups for a person to evaluate in the simulation experiment with discrete covariate described in section 5.1 of the paper.} 
\centering
\begin{tabular}{c c c c  c c c}
\hline\hline
Difficulty& \multicolumn{3}{c}{parameter combination}& Number of evaluations &Total number  & number of lineups\\
 \cline{2-4}
level & $n$ & $\sigma$ & $\beta$ &required($n_{\gamma}$) & of lineups & randomly shown \\
\hline
easy&100& 5&8 & 1& 12 & 1\\
&100&12&16 &1&&\\
&300& 5&5 &1&&\\
&300&12&10 &1 &&\\
\hline
medium&100& 5&3 &203 & 9 &2\\
&300& 5&2, 3 & 97, 1&&\\
\hline
hard&100&12&3, 8, 10 & 277, 126, 23& 18 &6\\
&300& 5&1 & 371 &&\\
&300&12&3, 5& 375, 74 &&\\
\hline
mixed&100& 5&1, 5, 0& 214, 2, 73 & 21 &1\\
&100&12&1& 100& &\\
&300& 5&0 & 73&&\\
&300&12&7, 1& 2, 152&&\\
\hline
Total &&&&&60&10\\
\hline
\end{tabular}
\label{tbl:dist_lineup1}
\end{table}

\section{Proof of the Lemma}

@@

\section{How much do null plots affect the choice?}

@@

We have seen that the subjects tend to pick the plot in the lineup that has the lowest $p$-value (Figure \ref{fig:P-val_log2}). What we are also interested in  is how this pick is affected by the distribution  of $p$-values of other plots in the lineup, particularly the $p$-value of the null plot with the strongest structure. If there is a null plot with a small $p$-value, or one close to that of the actual data plot, we would expect that subjects have a harder time detecting the actual data plot. Figure \ref{fig:pval_difference} investigates this. The difference between the $p$-value of the actual data is compared with the lowest from the null plots. This is plotted horizontally, and the proportion correct is plotted vertically. Negative values indicate lineups where the actual data plot had a smaller $p$-value than the minimum of the null plots. In experiment 1 (boxplots) there were a lot of lineups where the actual data plot had the smallest $p$-value, but only just. This caused quite some confusion for subjects, as seen because the variability in the proportion correct is huge for these lineups. Similarly large variability in correctness can be seen in the results of experiment 2 (scatterplots) except that the greater range of of differences in $p$-values shows the strength of subject's ability to pick the plot with most structure. Figure \ref{fig:P-val_log2} shed some more light on this story: when there is a big difference between the $p$-values (eg experiment 1, $\beta>7$) the subjects as one force chose the same plot. When there is less difference the distribution of counts is much more evenly spread between plots (eg experiment 1, $\beta=1$).

@@

\section{Type III error}~\label{sec:TypeIII}

@@

Figure \ref{fig:power_loess_effect} indicated that Type III error might be occurring in experiment 3: correct identification of the actual data plot is not positively associated with effect size. Teasing this out of the results is possible by looking at the reasons observers gave for their choices. Observers were provided with four possible reasons to use for their choice: (1) Most different plot, (2) Visible trend, (3) Clustering visible and (4) Other, with the option to use more than one. The task requested subjects to identify the plot that had the largest slope, which would correspond to choosing ``visible trend'' (2) as the reason for their choice. Reasons 1 or 3 would be indicative of Type III error. Figure \ref{fig:choice_reason} explores the reasons subjects gave for their choices. If there were no Type III errors committed, we would expect that people overwhelmingly using ``visible trend'' as their reason, or at least, when they use this reason they overwhelmingly correctly choose the actual data plot. This is not what we see. At left, are the reasons subjects gave for their choices --- 123 means that they gave all three reasons. The horizontal axis shows proportion of times that subjects correctly chose the actual data plot, and the reasons are sorted from most accurate to least accurate. The size of the point corresponds to the number of subjects putting this as the reason. Subjects that chose all three reasons almost always chose the actual data plot. This was followed by using 1 and 3, and then 1 and 4. The most common reasons given were reasons 1-3 individually, and the accuracy for these reasons ranged from 75\% for reason ``most different plot'' to 60\% for ``visible trend''.
At right is a simplified view, containing just the four possible reasons -- if the subject chose one of these, regardless if they also chose another reason it is counted. ``Visible trend'' comes in third. This is strong evidence that for many subjects even though they are correctly choosing the data plot, often they are cueing to other structure in the plot than the trend, making a Type III error.

@@

\section{Data Cleaning}

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.70}{\includegraphics{power_screening_subject.pdf}}
       \caption{The subject specific power estimated from mixed model for all the six screening criteria are shown. The corresponding power curves for conventional test is shown for comparison.}
       \label{fig:power_screening_subject}
\end{figure}

@@

>>>>

% for external reference
\usepackage{xr}
\externaldocument[p-]{paper_jasa}

@@

\title{Validation of Visual Statistical Inference, Applied to Linear Models (supplementary materials)}
%\author{Mahbubul Majumder, Heike Hofmann, Dianne Cook}
\date{\vspace{-.7in}}

@@

\maketitle

\vspace{1.5cm}

The materials in this document supplement the information presented in the manuscript ``Validation of Visual Statistical Inference, Applied to Linear Models".  Section \ref{sec:lineup_selection} describes how the lineups are presented to the subjects for evaluation, section \ref{sec:proof_lemma} presents the proof of the Lemma \ref{p-lemma} in the manuscript. The detailed discussion on how much the choice of null plots affect is included in Section \ref{sec:null_choice} which supplement the information in Section \ref{p-sec:null_choice} of the manuscript. Section \ref{sec:typeIII_error} includes the detailed discussion about the Type-III error supplementing Section  \ref{p-sec:typeIII_error}. The data cleaning process and how that may affect the result are discussed in Section \ref{sec:data_cleaning} which supplements Section \ref{p-sec:data_cleaning} of the manuscript.

\section{Selection of Lineups for each subject} \label{sec:lineup_selection}
Table \ref{tbl:dist_lineup1} shows how the lineups are presented to the subjects for evaluation for the simulation experiment with categorical covariate described in Section \ref{p-sec:category} of the manuscript . The lineups are divided into four groups such as easy, medium, hard and mixed based on the parameter combinations shown in the table. The number of lineups randomly chosen for each difficulty level is shown in the table as well. Note that every subject saw at least one easy lineup in the pool of 10 lineups shown. Some subjects saw more than one easy lineup that comes from the mixed group. The number of total evaluations required for each lineup is also shown in the table and a total of 300 subjects would provide a total of 3000 evaluations that should roughly satisfy that plan.

\begin{table}[hbtp]
%\renewcommand\thetable{}
\caption{Selection of 10 lineups for a person to evaluate in the simulation experiment with discrete covariate described in section \ref{p-sec:category} of the paper.} 
\centering
\begin{tabular}{c c c c  c c c}
\hline\hline
Difficulty& \multicolumn{3}{c}{parameter combination}& Number of evaluations &Total number  & number of lineups\\
 \cline{2-4}
level & $n$ & $\sigma$ & $\beta$ &required($n_{\gamma}$) & of lineups & randomly shown \\
\hline
easy&100& 5&8 & 1& 12 & 1\\
&100&12&16 &1&&\\
&300& 5&5 &1&&\\
&300&12&10 &1 &&\\
\hline
medium&100& 5&3 &203 & 9 &2\\
&300& 5&2, 3 & 97, 1&&\\
\hline
hard&100&12&3, 8, 10 & 277, 126, 23& 18 &6\\
&300& 5&1 & 371 &&\\
&300&12&3, 5& 375, 74 &&\\
\hline
mixed&100& 5&1, 5, 0& 214, 2, 73 & 21 &1\\
&100&12&1& 100& &\\
&300& 5&0 & 73&&\\
&300&12&7, 1& 2, 152&&\\
\hline
Total &&&&&60&10\\
\hline
\end{tabular}
\label{tbl:dist_lineup1}
\end{table}

\section{Proof of the Lemma} \label{sec:proof_lemma}

The proof of the Lemma \ref{p-lemma} in the manuscript is shown below;

@@

\section{How much do null plots affect the choice?} \label{sec:null_choice}

@@

We have seen that the subjects tend to pick the plot in the lineup that has the lowest $p$-value (Figure \ref{p-fig:P-val_log2} in the mauscript). What we are also interested in  is how this pick is affected by the distribution  of $p$-values of other plots in the lineup, particularly the $p$-value of the null plot with the strongest structure. If there is a null plot with a small $p$-value, or one close to that of the actual data plot, we would expect that subjects have a harder time detecting the actual data plot. Figure \ref{fig:pval_difference} investigates this. The difference between the $p$-value of the actual data is compared with the lowest from the null plots. This is plotted horizontally, and the proportion correct is plotted vertically. Negative values indicate lineups where the actual data plot had a smaller $p$-value than the minimum of the null plots. In experiment 1 (boxplots) there were a lot of lineups where the actual data plot had the smallest $p$-value, but only just. This caused quite some confusion for subjects, as seen because the variability in the proportion correct is huge for these lineups. Similarly large variability in correctness can be seen in the results of experiment 2 (scatterplots) except that the greater range of of differences in $p$-values shows the strength of subject's ability to pick the plot with most structure. Figure \ref{p-fig:P-val_log2} in the manuscript shed some more light on this story: when there is a big difference between the $p$-values (eg experiment 1, $\beta>7$) the subjects as one force chose the same plot. When there is less difference the distribution of counts is much more evenly spread between plots (eg experiment 1, $\beta=1$).

@@

\section{Type III error}~\label{sec:TypeIII}\label{sec:typeIII_error}

@@

Figure \ref{p-fig:power_loess_effect} in the manuscript indicated that Type III error might be occurring in experiment 3: correct identification of the actual data plot is not positively associated with effect size. Teasing this out of the results is possible by looking at the reasons observers gave for their choices. Observers were provided with four possible reasons to use for their choice: (1) Most different plot, (2) Visible trend, (3) Clustering visible and (4) Other, with the option to use more than one. The task requested subjects to identify the plot that had the largest slope, which would correspond to choosing ``visible trend'' (2) as the reason for their choice. Reasons 1 or 3 would be indicative of Type III error. Figure \ref{fig:choice_reason} explores the reasons subjects gave for their choices. If there were no Type III errors committed, we would expect that people overwhelmingly using ``visible trend'' as their reason, or at least, when they use this reason they overwhelmingly correctly choose the actual data plot. This is not what we see. At left, are the reasons subjects gave for their choices --- 123 means that they gave all three reasons. The horizontal axis shows proportion of times that subjects correctly chose the actual data plot, and the reasons are sorted from most accurate to least accurate. The size of the point corresponds to the number of subjects putting this as the reason. Subjects that chose all three reasons almost always chose the actual data plot. This was followed by using 1 and 3, and then 1 and 4. The most common reasons given were reasons 1-3 individually, and the accuracy for these reasons ranged from 75\% for reason ``most different plot'' to 60\% for ``visible trend''.
At right is a simplified view, containing just the four possible reasons -- if the subject chose one of these, regardless if they also chose another reason it is counted. ``Visible trend'' comes in third. This is strong evidence that for many subjects even though they are correctly choosing the data plot, often they are cueing to other structure in the plot than the trend, making a Type III error.

@@

\section{Data Cleaning}\label{sec:data_cleaning}

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.70}{\includegraphics{power_screening_subject.pdf}}
       \caption{The subject specific power ($K=1$) estimated from mixed model for all the six screening criteria are shown. The corresponding power curves for conventional test is shown for comparison.}
       \label{fig:power_screening_subject}
\end{figure}

@@

\bibliographystyle{asa}
\bibliography{references}
